How has the Brazilian Amazon been constructed as a problem? Presidential speeches and transnational politics since 1985

### 2023-02-03 Submission/Re-submission/draft/draft_resubmission.rmd
new text

---
title: How has the Brazilian Amazon been constructed as a problem? Presidential speeches and transnational politics since 1985
author:
- Livio Silva-Muller
- Henrique Sposito^[Both co-authors contributed equally to the article. Names are ordered alphabetically. Livio is a PhD Candidate in Anthropology and Sociology at the Geneva Graduate Institute and affiliated to the Albert Hirschman Centre on Democracy. Henrique is a PhD Candidate in International Relations and Political Sciences at the Geneva Graduate Institute and affiliated to the Centre for International Environmental Studies.]
date: 27 January 2023
output:
  pdf_document: default
  latex_engine: xelatex
  fig_caption: yes
  html_document:
    df_print: paged
fontsize: 12pt
mainfont: Times New Roman
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage[T1]{fontenc}
   - \usepackage[utf8]{inputenc}
   - |
    \providecommand{\keywords}[1]{\textbf{Keywords:} #1}
    \providecommand{\wordcount}[1]{\textbf{Word Count:}#1}
always_allow_html: TRUE
bibliography: references.bib
csl: chicago.csl
abstract: blank
---

@@

Each of the above statements contains an implicit or explicit assumption of what needs to be solved, or in other words, it constructs the Amazon as a particular problem. Scholars studying the Brazilian Amazon often identify different governments throughout Brazilian history as proponents of a view of the region, its people, and forest [@drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021]. While the military dictatorship (1964-85) is, for example, associated with understanding the Amazon as issues of national sovereignty and economic integration, Sarney's (1985-1990) and Lula's (2003-2010) presidencies are often tied to environmental conservation. Yet, such monolithic representations advance the view that governments understood and acted consistently towards the Amazon as a specific problem, masking government diversity of discursive constructions and policies.

In this article, we show how the Brazilian Amazon has been constructed as a problem in presidential discourses since 1985. By adapting @bacchi2009's theory of problem-representation and @keck1998's theory of transnationalism (1998), we offer a framework that (1) explains transnational variation in problem-constructions as a function of speech setting and policies' outcomes, and (2) focuses on discourses rather than policies because they allow us to capture the diversity of ways policy goals are constructed to different audiences.

Many analyses of Brazilian environmental policies approach it as a national problem or an international problem, without a discussion of how important national actors behave differently in both scenarios. Our framework and methodology identify and accounts for this variation. We show, for instance, that when speaking to farmers in Cuiaba, one of Brazil's agricultural powerhouses, in a moment when deforestation is increasing, it is likely that any president would sidestep environmental issues, as President Lula did in the opening epigraph. Meanwhile, even President Bolsonaro at the United Nations General Assembly in New York City, spoke about protecting the Amazonian forests and its peoples.

Our focus on presidential speeches affords us analytical leverage and it is crucial to understand policy. The Brazilian executive branch of government, led by the president, has historically been the primary actor in shaping public policy in Brazil (Pinto 2016). While presidential speeches in Brazil have been studied for topics such as race relations [@silva2019] and inequality [@grangeia2017], the environment remains largely absent. Presidential speeches legitimize perceptions about the Amazon that are key for policy adoption and implementation [see @alesina2009; @lopez2020]. This is especially pertinent for deforestation in the Amazon as expectations of local actors, generated from material and discursive governmental practices, are crucial in decisions to deforest on the ground [@assuncao2015; @capobianco2019; @capobianco2021; @campbell2015].

Our findings are threefold. First, we find that the frequency in which the Amazon appears as a topic in presidential speeches is driven by domestic and international environmentally related events. Second, the same government adopts a diverse range of problem-construction. While the constructions of the Amazon as an issue of economic integration were more salient from 1985 to the mid-2000s, environmental conservation and social development problem-constructions steadily grew since the 2000s, temporarily surpassing economic integration as a more salient construction from 2010 to 2015. Constructing the Amazon as an issue of sovereignty steadily increased after 2010. Third, we find that presidents are more likely to construct the Amazon as an issue of environmental conservation when they speak at non-Amazonian countries outside of Brazil. Within the Amazon region and neighboring countries, presidents tend to construct the Amazon as an issue of social development and economic integration. This article provides the first comprehensive overview of when, how, and where the Amazon is constructed as problems in presidential speeches.

@@

@@

We know, however, that there is more variation within a specific government policy, especially if we consider the Amazon a policy object demanding dedicated policy attention. For instance, while many argue that environmental activism in Brazil emerged despite the military dictatorship [@hochstetler2000; @wood1993], @rajao2021 demonstrates how the strengthening of the Forest Code in 1964 was also a military priority. Furthermore, a substantial part of Lula's administration (2003-2010) effectiveness in reducing deforestation in the Amazon was achieved because of deforestation leakage to other biomes (Garret, 2022). At the same time, Lula's administration (2003-2010) provided record high subsidies to the expansion of agriculture in Brazil [@capobianco2021]. That is, we must look at the Amazon beyond the environmental domain as a forest, region, and peoples.

@@

The possibility of different problem-constructions for the same policy object in political discourse, requires an explanation of why they vary. Many analyses of Brazilian policies towards the Amazon approach it as a national problem or an international problem, without a discussion of how important national actors behave differently in both scenarios. By focusing on problem-constructions in presidential speeches, with an analytical strategy that distinguishes the location of the speech, we can move the empirical debate on the transnational nature of the Amazon forward. There are two main contenders for tying the problem-construction to international politics: Putnam's two-level game and Keck and Sikkink's transnational networks.

@@

To test our conceptual framework, we need to operationalize our main dependent variable, these are the possible problem-constructions for a policy object. To identify and classify Amazonian speeches as specific types of problem constructions, we developed a codebook inductively and with assistance of the literature. We identify three historical-political constructions: national sovereignty, economic integration, and environmental conservation. We argue a fourth separate one also exists: social development. The social development construction focuses on the constitutional rights and dignity of the inhabitants of the Amazonian region.

@@

Our explanation for variation in problem-constructions is based on first the setting of a speech and second the policy outcomes.

@@

First, speeches can take place in Amazonian States inside Brazil. As a constituency with likely distinct demands than in other Brazilian states, presidents can try to communicate with them highlighting what presidents project their demands to be. Second, and consequently, Brazilian states outside the Amazon are another setting. In this case, as populations with some stake in the Amazon but not residing there, they might have different views of the Amazon.\

@@

While we would ideally control for the exact venue of the speech (multilateral mechanism, private fora, etc.), so we could have an even more precise identification of the audience, the location information of the speeches does not allow so. Nevertheless, we believe these four settings and their different audiences are already enough to demonstrate whether and how problem-construction indeed changes.

@@

Many scholars inquire and find causal relationships between discourse and policy outcomes. While our research design and data does not allow us to test so, we do make arguments that might suggest a direction in certain parts....[further develop this limitation]

@@

As a dataset like this was never constructed before, we start with longitudinal descriptive statistics, which can reveal interesting trends. First, we identify years in which Amazon as a general topic has been discussed more frequently than others and possible drivers. To do so, we take the share of yearly speeches that mention Amazon independently of the specific problem construction. Next, we discuss the evolution of yearly problem-constructions in time by showing our data longitudinally.We conclude with four separate fixed-effect logistic regression models that predict the probability of each problem-construction relative to the setting in which a speech takes place and the policy outcome.

The dependent variable of each model is the share of one pure-type problem construction against all other three problem constructions. We use four separate logistic regressions to provide unambiguous interpretations and comparisons, as a multinomial model would entail four categories on both dependent and independent variables. We include presidential fixed effects because we can expect individual-specific characteristics (i.e. presidential ideology) to have an effect on the outcome. If the president's political ideology is a good predictor of pure-type variation, it could indicate that the president is rather isolated from settings and that they can follow their own worldviews[^2]. Fixed-effects control for this issue[@allison2009fixed], and other unit-unvarying characteristics. Our main explanatory variables, setting and policy outcomes, are included as described in section 3.1.2.  We add a third variable, which takes the value of 1 if the saliency of the Amazon as a general topic was above average that year. We do so to test peaks in the Amazon as a general topic driving any particular problem-construction. Finally, we add two standard controls: yearly inflation rate (numerical) and election year (dummy).

@@

```{r setting vs outcome, fig.cap="Problem-constructions, settings, and policy outcomes", echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, include=TRUE}
amazon_def_year <- amazon_def_year %>%
  mutate(extremes= ifelse(year == 2004|year == 2012, "yes","no"))
ama_mx %>%
  count(mx_cat, location_cat, year) %>%
  group_by(year) %>%
  mutate(perc = n / sum(n)) %>% 
  filter(location_cat == "Amazonian States"|
           location_cat == "Non Amazonian Countries",
         mx_cat == "Pure Economic Integration"|
           mx_cat == "Pure Environmental Conservation") %>%
  ggplot() +
  geom_smooth(aes(x=as.numeric(year), y = perc, color=mx_cat),
              se = FALSE, linewidth =0.75, linetype="solid") +
  scale_y_continuous(labels = percent_format(accuracy = 1))+
  facet_wrap(~location_cat+mx_cat)+
  labs(x = "", y = "", title ="Relationship between problem-constructions and deforestation rates",
       subtitle = "Bars indicated yearly deforestation level and curves indicated yearly share of problem-construction.",
       caption = "Curves in the plot were smoothed using loess method.")+
  theme_am()+
  geom_col(amazon_def_year, mapping = aes(x = year, y = def_year/120,fill=extremes),alpha=.5)+
  scale_fill_manual(values=c("lightgrey", "darkolivegreen"))+
  annotate("text", x = 2009, y = 0.20, label = "27'772 km2\ndeforested" , size = 1.5, color="darkolivegreen", face="bold")+
  annotate("text", x = 2015, y = 0.038, label = "4'571 km2\ndeforested", size = 1.5, color="darkolivegreen", face="bold")
```

>>>>

---
title: How has the Brazilian Amazon been constructed as a problem? Presidential speeches and transnational politics since 1985
author:
- Livio Silva-Muller
- Henrique Sposito^[Both co-authors contributed equally to the article. Names are ordered alphabetically. Livio is a PhD Candidate in Anthropology and Sociology at the Geneva Graduate Institute and affiliated to the Albert Hirschman Centre on Democracy. Henrique is a PhD Candidate in International Relations and Political Sciences at the Geneva Graduate Institute and affiliated to the Centre for International Environmental Studies.]
date: 2 February 2023
output:
  pdf_document: default
  latex_engine: xelatex
  fig_caption: yes
  html_document:
    df_print: paged
fontsize: 12pt
mainfont: Times New Roman
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage[T1]{fontenc}
   - \usepackage[utf8]{inputenc}
   - |
    \providecommand{\keywords}[1]{\textbf{Keywords:} #1}
    \providecommand{\wordcount}[1]{\textbf{Word Count:}#1}
always_allow_html: TRUE
bibliography: references.bib
csl: chicago.csl
abstract: blank
---

@@

Each of the above statements contains an implicit or explicit assumption of what needs to be solved, or in other words, it constructs the Amazon as a particular problem. Scholars studying the Brazilian Amazon often identify different governments throughout Brazilian history as proponents of a view of the region, its people, and forest [@drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021]. While the military dictatorship (1964-85) and Bolsonaro's presidency (2019-2022) are, for example, associated with understanding the Amazon as issues of national sovereignty and economic integration, Sarney's (1985-1990) and Lula's (2003-2010) presidencies are often tied to environmental conservation. Yet, such  representations advance the view that governments understood and acted consistently towards the Amazon as a specific problem, masking government diversity of discursive constructions and policies.

In this article, we show how the Brazilian Amazon has been constructed as a problem in presidential discourses since 1985. By combining @bacchi2009's theory of problem-representation and @keck1998's theory of transnationalism (1998), we offer a framework that (1) explains transnational variation in problem-constructions depending on speech setting and policies' outcomes, and (2) focuses on discourses rather than policies because they allow us to capture the diversity of ways policy goals are constructed to different audiences.

Many analyses of Brazilian environmental policies approach it as a national or an international environmental problem, without a discussion of how important national actors behave differently in both scenarios. Our framework and methodology identify and accounts for this variation. We show, for instance, that when speaking to farmers in Cuiaba, one of Brazil's agricultural powerhouses, in a moment when deforestation is increasing, it is likely that any president would sidestep environmental issues, as President Lula did in the opening epigraph. Meanwhile, even President Bolsonaro at the United Nations General Assembly in New York City, spoke about protecting the Amazonian forests and its peoples.

Our focus on presidential speeches affords us analytical leverage and it is crucial to understand policy. The Brazilian executive branch of government, led by the president, has historically been the primary actor in shaping public policy in Brazil(Pinto 2016). While presidential speeches in Brazil have been studied for topics such as race relations [@silva2019] and inequality [@grangeia2017], the Amazon forest and its peoples remains largely absent. Presidential speeches legitimize perceptions about the Amazon that are key for policy adoption and implementation [see @alesina2009; @lopez2020]. This is especially pertinent for deforestation in the Amazon as expectations of local actors, generated from material and discursive governmental practices, are crucial in decisions to deforest on the ground [@assuncao2015; @capobianco2019; @capobianco2021; @campbell2015].

Our findings are threefold. First, we find that the frequency in which the Amazon appears as a topic in presidential speeches is driven by domestic and international coverage of environmentally related events. Second, while the same government adopts a diverse range of problem-construction, we identify a process of convergence:  the constructions of the Amazon as an issue of economic integration were more salient from 1985 to the mid-2000s, but environmental conservation and social development problem-constructions steadily grew to temporarily surpass economic integration as from 2010 to 2015. Constructing the Amazon as an issue of sovereignty steadily increased after 2008, to the extent that all problem-constructions converge to being deployed in a balanced way after 2015. Third, we find a cumulative effect between speech setting and policy outcome. Presidents are more likely to construct the Amazon as an issue of environmental conservation when they speak at non-Amazonian countries outside of Brazil and deforestation figures are positive. Within the Amazon region and neighboring countries, presidents tend to construct the Amazon as an issue of social development and economic integration. When deforestation figures are negative, presidents tend to side-step environmental conservation. Overall...

@@

We divide our theoretical framework in two parts. We start with an explanation of what are problem-constructions, our dependent variable at large, and why presidential discourse are an interesting instance to identify them. Next, we argue that problem-constructions can vary depending on the transnational setting of the speech and on how good policy outcomes are in the moment of speech, our main independent variables.

@@

@@

 Many analyses of Brazilian policies towards the Amazon approach it as a national problem or an international problem, without a discussion of how important national actors behave differently in both scenarios. The possibility of different problem-constructions for the same policy object in political discourse, requires an explanation of why they vary. By focusing on problem-constructions in presidential speeches, with an analytical strategy that distinguishes the location of the speech, we can move the empirical debate on the transnational nature of the Amazon forward. There are two main contenders for tying the problem-construction to international politics: Putnam's two-level game and Keck and Sikkink's transnational networks.

@@

To test our conceptual framework, we need to operationalize our main dependent variable, the possible problem-constructions for a policy object. To identify and classify Amazonian speeches as specific types of problem constructions, we developed a codebook inductively and with assistance of the literature. We identify three historical-political constructions: national sovereignty, economic integration, and environmental conservation. We argue a fourth separate one also exists: social development. The social development construction focuses on the constitutional rights and dignity of the inhabitants of the Amazonian region.

@@

Our explanation for variation in problem-constructions is based on first the setting of a speech and second the policy outcomes. This comprise, at large, our main independent variables.

@@

First, speeches can take place in Amazonian States inside Brazil. As a constituency with likely distinct demands than in other Brazilian states, presidents can try to communicate with them highlighting what presidents project their demands to be. Second, and consequently, Brazilian states outside the Amazon are another setting. In this case, as populations with some stake in the Amazon but not residing there, they might have different views of the Amazon.

@@

While we would ideally control for the exact venue of the speech (multilateral mechanism, private fora, etc.), so we could have an even more precise identification of the audience, the location information of the speeches does not allow so. Nevertheless, we believe these four settings and their different audiences are already enough to demonstrate whether and how problem-construction indeed changes transnationally.

@@

Many scholars inquire and find causal relationships between discourse and policy outcomes. Our research design and data does not allow us to test so.  While we might suggest a particular direction in certain cases, none of our claims are linearly causal.

@@

As a dataset like this was never constructed before, we start with longitudinal descriptive statistics, which  reveals interesting trends. First, we identify years in which Amazon as a general topic has been discussed more frequently and possible drivers. To do so, we take the share of yearly speeches that mention Amazon independently of the specific problem construction. Next, we discuss the evolution of yearly problem-constructions in time by showing our data longitudinally.We conclude with four separate fixed-effect logistic regression models that predict the probability of each problem-construction relative to the setting in which a speech takes place and the policy outcome.

The dependent variable of each model is the share of one pure-type problem construction against all other three problem constructions. We use four separate logistic regressions to provide unambiguous interpretations and comparisons, as a multinomial model would entail four categories on both dependent and independent variables. We include presidential fixed effects because we can expect individual-specific characteristics (i.e. presidential ideology) to have an effect on the outcome. If the president's political ideology is a good predictor of pure-type variation, it could indicate that the president is rather isolated from settings and that they can follow their own worldviews[^2]. Fixed-effects control for this issue[@allison2009fixed], and other unit-unvarying characteristics. Our main explanatory variables, setting and policy outcomes, are included as described in section 3.1.2. We add a third variable, which takes the value of 1 if the salience of the Amazon as a general topic was above average that year. We do so to test peaks in the Amazon as a general topic driving any particular problem-construction. Finally, we add two standard controls: yearly inflation rate (numerical) and election year (dummy).

@@

```{r setting vs outcome, fig.cap="Problem-constructions, settings, and policy outcomes", echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, include=TRUE}
amazon_def_year <- amazon_def_year %>%
  mutate(extremes= ifelse(year == 2004|year == 2012, "yes","no"))
ama_mx %>%
  count(mx_cat, location_cat, year) %>%
  group_by(year) %>%
  mutate(perc = n / sum(n)) %>% 
  filter(location_cat == "Amazonian States"|
           location_cat == "Non Amazonian Countries",
         mx_cat == "Pure Economic Integration"|
           mx_cat == "Pure Environmental Conservation"|
           mx_cat == "Pure Social Development"|
           mx_cat == "Pure National Sovereignty") %>%
  ggplot() +
  geom_smooth(aes(x=as.numeric(year), y = perc, color=location_cat),
              se = FALSE, linewidth =0.75, linetype="solid") +
  scale_y_continuous(labels = percent_format(accuracy = 1))+
  facet_wrap(~mx_cat)+
  labs(x = "", y = "", title ="Relationship between problem-constructions and deforestation rates",
       subtitle = "Bars indicated yearly deforestation level and curves indicated yearly share of problem-construction.",
       caption = "Curves in the plot were smoothed using loess method.")+
  theme_am()+
  geom_col(amazon_def_year, mapping = aes(x = year, y = def_year/120,fill=extremes),alpha=.5)+
  scale_fill_manual(values=c("lightgrey", "darkolivegreen"))+
  annotate("text", x = 2009, y = 0.20, label = "27'772 km2\ndeforested" , size = 1.5, color="darkolivegreen", face="bold")+
  annotate("text", x = 2015, y = 0.038, label = "4'571 km2\ndeforested", size = 1.5, color="darkolivegreen", face="bold")
```
